---
title: "Marin County RIPA Stop Analysis"
author: "Weiyang Wang"
date: "2026-05-29"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r message=FALSE, warning=FALSE}
{knitr::opts_chunk$set(echo = TRUE)} 
library(readxl) 
library(knitr)
library(tidyverse) 
RIPA_MARIN_2024 = read_excel("Final files/RIPA Stop Data_Marin 2024_final.xlsx")
```

# Racial Justice Act and RIPA

## Introduction

Under Penal Code § 745, The state shall not seek or obtain a criminal conviction or seek, obtain, or impose a sentence on the basis of race, ethnicity, or national origin. A violation is established if the defendant proves, by a preponderance of the evidence, any of the following:

(3) The defendant was charged or convicted of a more serious offense than defendants of other races, ethnicities, or national origins who have engaged in similar conduct and are similarly situated, and the evidence establishes that the prosecution more frequently sought or obtained convictions for more serious offenses against people who share the defendant’s race, ethnicity, or national origin in the county where the convictions were sought or obtained.

Law enforcement interactions are intricately tied to RJA claims. Police reports, arrest charges, and evidentiary citations form the foundation of prosecutorial charging decisions. If law enforcement disproportionately arrests or cites specific racial/ethnic groups, it directly feeds into the systemic disparities challenged under an (a)(3) claim.

A defendant makes a prima facie showing by producing facts that, if true, establish a substantial likelihood of racial discrimination. Once that threshold is met, the court must hold an evidentiary hearing. Statistical evidence and aggregate data are expressly authorized forms of proof.

To make a prima facie showing with the available RIPA data, we need to analyze how (1) race, ethnicities, or national origins, (3) resulted in prosecution sought or obtained convictions for more serious offense (2) on a similar conduct and are similarly situated.

This report will:

(1) Use race and language proficiency as nexuses for national origins; and

(2) Condition on the type of stop and the reason for the stop to ensure that the defendant’s conduct leading to the stop is comparable to that of others. Geographic comparability will be addressed by conditioning on the law enforcement agency; and

(3) Focus on the results of the stop, specifically citations and arrests that initiate criminal proceedings. These outcomes are relevant because they create the opportunity for prosecution, whereas non-enforcement outcomes, such as verbal warnings, do not.

With only RIPA data available, focusing on law enforcement actions that could initiate criminal proceedings sheds light on potential RJA violations, as these actions form the critical gateway through which disparities in enforcement may translate into prosecutorial decisions and, ultimately, charging and conviction outcomes.

---

# Data sets

RIPA stop data are from <https://openjustice.doj.ca.gov/data>

## Description of Each Column as Basis for RJA(a)(3) Motion

Description of each field of data can be found here: <https://data-openjustice.doj.ca.gov/sites/default/files/dataset/2025-12/ripa-stop-dataset-readme-2024.pdf>

Some of the important ones are listed here:

- `AGENCY_NAME` is the name of the law enforcement agency, which includes

```{r}
agency_name_list = unique(RIPA_MARIN_2024$AGENCY_NAME)
agency_name_list
```

- `TIME_OF_STOP` is the time of stop, in 24 hour display HH:MM:SS

- `STOP_DURATION` is the duration of stop, in minutes

- `RAE_FULL` is the Perceived Race or Ethnicity of the Person stopped, identified as
  - 1: Asian or `RAE_ASIAN`
  - 2: Black/African American or `RAE_BLACK_AFRICAN_AMERICAN`
  - 3: Hispanic/Latine(x) or `RAE_HISPANIC_LATINEX`
  - 4: Middle Eastern/ South Asian or `RAE_MIDDLE_EASTERN_SOUTH_ASIAN`
  - 5: Native American or `RAE_NATIVE_AMERICAN`
  - 6: Pacific Islander or `RAE_PACIFIC_ISLANDER`
  - 7: White or `RAE_WHITE`
  - 8: Multiracial or `RAE_MULTIRACIAL`

- `G_FULL` is the Perceived gender of person stopped, identified as
  - 1: Cisgender Man/Boy or `G_CISGENDER_MAN`
  - 2: Cisgender Woman/Girl or `G_CISGENDER_WOMAN`
  - 3: Transgender Man/Boy or `G_TRANSGENDER_MAN`
  - 4: Transgender Woman/Girl or `G_TRANSGENDER_WOMAN`
  - 5: Nonbinary Person or `G_NONBINARY_PERSON`

- `AGE` is the Perceived age of person stopped, in numeric, which can be further categorized into `AGE_GROUP`:
  - 1: 1-9 years
  - 2: 10-14 years
  - 3: 15-17 years
  - 4: 18-24 years
  - 5: 25-34 years
  - 6: 35-44 years
  - 7: 45-54 years
  - 8: 55-64 years
  - 9: 65 years and older

- `LIMITED_ENGLISH_FLUENCY` is the Person's English fluency, with 1 being limited or no fluency.

- `PERSON_UNHOUSED` is the The stopped person perceived to be unhoused, with 1 being yes.

### Demographic Reporting Method

Demographic information pertaining to individual identity is based on officer perception rather than self-identification, and includes race/ethnicity, gender, age, sexual orientation, English fluency, disability status, and housing status.

## Type of Stop and Reason of Stop as a Showing of "Similarly Situated"

For the purposes of establishing RJA (a)(3) violation, the following variable will be taken into consideration:

- `TOS_VEHICULAR` as Type of Stop: Vehicular, with a value of 1 being yes.
- `TOS_BICYCLE` as Type of Stop: Bicycle, with a value of 1 being yes.
- `TOS_PEDESTRIAN` as Type of Stop: Pedestrian (all other stops, e.g. passengers on bus or train, skateboarders, etc.,) with a value of 1 being yes.

Welfare Check and Call for Services are excluded from consideration, as the general public are not subject to RJA, and law enforcement agency was the responder to those inquires.

```{r}
RIPA_MARIN_2024 = RIPA_MARIN_2024|>
  filter(TOS_VEHICULAR==1 | TOS_BICYCLE==1 | TOS_PEDESTRIAN==1)
```

The relevant reason for stop (`REASON_FOR_STOP`) are divided into the following categories:

- 1: Traffic Violation
- 2: Reasonable suspicion the person was engaged in criminal activity
- 3: Parole/probation/PRCS/mandatory supervision
- 9: Probable cause to arrest or search

```{r}
rfs_list = tibble(
  c(1, 2, 3, 9),
  c(
    "Traffic Violation",
    "Reasonable suspicion the person was engaged in criminal activity",
    "Parole/probation/PRCS/mandatory supervision",
    "Probable cause to arrest or search"
    )
)
```

Other stop-reason categories, such as student-related stops and consensual searches, are omitted because they are less relevant to establishing a prima facie violation under Penal Code § 745(a)(3). Likewise, more detailed classifications of reasonable suspicion and probable cause are not analyzed at this stage.

## Outcome of the Stop

RJA (a)(3) focuses on whether the prosecution "more frequently sought or obtained convictions for more serious offenses." Criminal proceedings arising from a stop typically begin with either a citation or an arrest. Accordingly, for purposes of analyzing potential disparities using RIPA data, this report focuses on stop outcomes that may initiate or expose an individual to criminal proceedings.

**Arrest**

- `ROS_CUSTODIAL_WITHOUT_WARRANT`: Result of stop: custodial arrest without warrant
- `ROS_CUSTODIAL_WARRANT`: Result of stop: custodial pursuant to outstanding warrant
  - This report will omit custodial pursuant to outstanding warrant, as this is race neutral.
  
**Citation**

- `ROS_CITATION`: Result of stop: citation for infraction
  - This report will omit citation for infraction, as this is usually traffic tickets
- `ROS_IN_FIELD_CITE_RELEASE`: Result of stop: in field cite and release

**Search (could include more depends on probative value)**

- `NFA_SEARCH_PERSON`: Non-force action: Search of person was conducted
- `NFA_SEARCH_PROPERTY`: Non-force action: Search of property was conducted

**Use of Force (could include more depends on probative value)**

- `OFA_HANDCUFFED`: Force action: Handcuffed or flex cuffed
- `OFA_PHYSICAL_COMPLIANCE`Force action: Physical compliance tactics and techniques
- `OFA_FIREARM_POINT`: Force action: Firearm pointed at person
- `OFA_FIREARM_DISCHARGE`: Force action: Firearm discharged

**No action**

- `ROS_NO_ACTION`: Result of stop: no action

## Limitations on Data

A relatively small amount of data with errors that should have been identified and corrected reached completed status. These known errors are listed below. Each count reflects the number of records, across all agencies that submitted 2024 data, containing the respective error.

Specific error to Marin County are not identified. Some worth noting errors and ways to address these minors are listed below:

- School-Related Reporting Issues
  - **27 cases**: Person identified as a K–12 student, but stop not marked as occurring on school grounds.
  - **59 cases**: Stop marked as occurring on K–12 grounds, but school name missing.
  - **114 cases**: Stop on K–12 grounds, but invalid school code provided.
  - **2 cases**: Education Code stop reason selected, but no discipline code provided.
  - **19 cases**: Non-student associated with a student-related stop reason.
  - **1 case**: Written statement obtained, but stopped person was not a student.
- Stop Reason and Offense Code Inconsistencies
  - **18 cases**: Traffic violation selected, but traffic violation type missing.
  - **34 cases**: Traffic violation selected, but CJIS offense code missing.
  - **15 cases**: Traffic violation selected, but reasonable suspicion fields completed instead.
  - **35 cases**: Reasonable suspicion selected, but supporting subcategories missing.
  - **3 cases**: Probable cause offense code entered when stop reason was not probable cause.
  - **27 cases**: Reasonable suspicion offense code entered when stop reason was not reasonable suspicion.
  - **8 cases**: Traffic violation offense code entered when stop reason was not traffic violation.
- Search-Related Issues
  - **106 cases**: Search conducted, but basis for search not documented.
  - **50 cases**: Search basis documented, but no search indicated.
  - **34 cases**: Consensual encounter resulting in search selected, but no search action recorded.
  - **5 cases**: Search based on consent, but consent action not documented.
  - **293 cases**: Consent to search given, but consent type not specified.
  - **1 case**: Vehicle inventory search basis selected, but no property search recorded.
- Officer Action Data Missing
  - **677 cases**: Missing non-force action information.
    - This report will omit entry with missing non-force action information, when appropriate.
  - **1,279 cases**: Missing force action information.
    - This report will omit entry with missing force action, when appropriate.
- Contraband and Evidence Issues
  - **1 case**: Contraband/evidence marked as not discovered, but a contraband/evidence subtype was selected.
- Property Seizure Issues
  - **267 cases**: Property seized, but no basis for seizure documented.
  - **31 cases**: Seizure based on contraband, but no contraband discovered.
  - **33 cases**: Seizure based on evidence, but no evidence discovered.
  - **284 cases**: Property seized, but property type not documented.
  - **18 cases**: No property seized, but seizure basis entered.
  - **25 cases**: No property seized, but property type entered.
- Stop Outcome Issues
  - **300 cases**: Missing stop result/outcome information.
    - Missing stop result are not included in the sample set.
- Vehicle Stop Issues
  - **341 cases**: Passenger identification/run recorded, but stop was not classified as a vehicle stop.

---

# Marin County Wide Report

## Race and Ethnicities

**Data Control and Processing**: This analysis uses 2024 California Racial and Identity Profiling Act (RIPA) data from Marin County to examine whether law enforcement outcomes vary across racial and ethnic groups among individuals stopped for similar reasons. To promote comparability, the analysis is conducted separately for each of the following stop-reason categories (`REASON_FOR_STOP`):

- 1: Traffic Violation
- 2: Reasonable suspicion the person was engaged in criminal activity
- 3: Parole/probation/PRCS/mandatory supervision
- 9: Probable cause to arrest or search

For each stop-reason category, race and ethnicity are identified using the `RAE_FULL` variable and recoded into the following groups: Asian, Black, Hispanic, Middle Eastern/South Asian, Native American, Pacific Islander, White, and Multiracial. Records with missing race or ethnicity information are excluded from the analysis.

Each stop is converted into a series of binary indicators representing key enforcement outcomes. 
- Arrests are defined as custodial arrests without a warrant (`ROS_CUSTODIAL_WITHOUT_WARRANT == 1`).
- Citations are defined as in-field cite-and-release dispositions (`ROS_IN_FIELD_CITE_RELEASE == 1`). 
- Searches are defined as any search of a person or property (`NFA_SEARCH_PERSON == 1` or `NFA_SEARCH_PROPERTY == 1`). 
- Use of force is defined as any instance of handcuffing, physical compliance techniques, firearm pointing, or firearm discharge (`OFA_HANDCUFFED`, `OFA_PHYSICAL_COMPLIANCE`, `OFA_FIREARM_POINT`, or `OFA_FIREARM_DISCHARGE`). 
- Stops resulting in no enforcement action are identified by `ROS_NO_ACTION == 1`.

Within each stop-reason category, the data are grouped by race and ethnicity. For each group, the analysis calculates the total number of stops, the total number of arrests, citations, searches, uses of force, and no-action outcomes, as well as the corresponding rates (means of binary indicators) for each outcome. Rates are calculated as the proportion of stops within a racial or ethnic group that resulted in the specified outcome.

**Descriptive Outcome**

To reduce the influence of small sample sizes and unstable estimates, racial and ethnic groups with fewer than ten observations within a stop-reason category are excluded from the summary. Comparative tables are produced only when at least two racial or ethnic groups meet this threshold. The resulting summaries provide descriptive measures of enforcement outcomes across racial and ethnic groups for individuals stopped under comparable circumstances.

**Statistical Analysis: Odds Ratio and 95% Confidence Interval**

To evaluate potential disparities in law enforcement outcomes across racial and ethnic groups, we calculated odds ratios (ORs) and 95% confidence intervals (CIs) for key enforcement outcomes, results, including arrests, citations, searches, use of force, and no-action outcomes. 

The odds ratio (OR) is a standard measure used in statistical analysis to compare the likelihood of an outcome between two groups. It expresses how the odds of a specific event in one group relate to the odds of the same event in a reference group. An odds ratio of 1 indicates equal odds between groups, meaning no difference in likelihood. An odds ratio greater than 1 indicates higher odds of the outcome in the group of interest relative to the reference group, while an odds ratio less than 1 indicates lower odds. Unlike raw percentages, odds ratios provide a relative measure of association that is symmetric and comparable across groups and outcomes.

**Methodology**: 

The association between exposure and outcome was quantified using the **odds ratio (OR)** derived from a 2×2 contingency table.

|               | Outcome Present | Outcome Absent |
|---------------|----------------|----------------|
| Test          | a              | b              |
| Compare       | c              | d              |

The odds ratio was calculated as:

$$
OR = \frac{a/b}{c/d} = \frac{ad}{bc}
$$

An OR greater than 1 indicates higher odds of the outcome among exposed individuals, an OR less than 1 indicates lower odds, and an OR equal to 1 indicates no association.

The 95% confidence interval (CI) for the odds ratio was calculated using the logarithmic transformation method. The natural logarithm of the odds ratio was obtained as:

$$
\ln(OR) = \ln\left(\frac{ad}{bc}\right)
$$

The standard error of the log odds ratio was calculated as:

$$
SE\left[\ln(OR)\right] =
\sqrt{\frac{1}{a}+\frac{1}{b}+\frac{1}{c}+\frac{1}{d}}
$$

The confidence limits on the logarithmic scale were computed as:

$$
\ln(OR) \pm 1.96 \times SE\left[\ln(OR)\right]
$$

The lower and upper confidence limits were then exponentiated to obtain the 95% confidence interval:

$$
95\%\,CI =
\left(
e^{\ln(OR)-1.96\times SE},
\;
e^{\ln(OR)+1.96\times SE}
\right)
$$

```{r}

# Function to calculate Odds Ratio and 95% CI
calc_or_ci = function(summary_df,
                      outcome_total_col,
                      group_col,
                      reference_group,
                      outcome_name = "Outcome") {
  
  ref_row = summary_df[summary_df[[group_col]] == reference_group, ]
  
  if (nrow(ref_row) == 0) {
    stop(paste("Reference group", reference_group, "not found"))
  }
  
  ref_events = ref_row[[outcome_total_col]]
  ref_nonevents = ref_row$n - ref_events
  
  results = data.frame(
    Group = character(),
    "Odd Ratio" = numeric(),
    CI_lower = numeric(),
    CI_upper = numeric(),
    Outcome = character(),
    stringsAsFactors = FALSE
  )
  
  for (i in 1:nrow(summary_df)) {
    
    row = summary_df[i, ]
    group_name = row[[group_col]]
    
    events = row[[outcome_total_col]]
    nonevents = row$n - events
    
    odd_ratio = NA
    lower = NA
    upper = NA

    if (events > 0 && nonevents > 0 && ref_events > 0 && ref_nonevents > 0) {
      
      odd_ratio = (events / nonevents) / (ref_events / ref_nonevents)
      
      se = sqrt(
        1/events +
        1/nonevents +
        1/ref_events +
        1/ref_nonevents
      )
      
      lower = exp(log(odd_ratio) - 1.96 * se)
      upper = exp(log(odd_ratio) + 1.96 * se)
    }
    
    results = rbind(
      results,
      data.frame(
        Group = group_name,
        "Odd Ratio" = odd_ratio,
        CI_lower = lower,
        CI_upper = upper,
        Outcome = outcome_name,
        stringsAsFactors = FALSE
      )
    )
  }
  return(results)
}

```

Within the context of this code, the resulting odds ratios will show:

- The relative likelihood that individuals from each racial or ethnic group experience a given enforcement outcome compared to White individuals, conditional on being stopped.
- Whether observed differences are likely to reflect more than random variation in the sample, as indicated by whether the 95% confidence interval includes 1.
- How these relative differences vary across different enforcement outcomes (arrest, citation, search, force, and no-action), allowing for comparison of patterns across types of police action.

Importantly, the output of this analysis is *descriptive and associative, not causal*. It provides prima facie statistical evidence of differential treatment that may warrant further investigation in the context of prosecutorial and enforcement decision-making.

```{r, results = 'asis'}

for (i in 1:dim(rfs_list)[1]) {
  rfs = as.numeric(rfs_list[i,1])
  
  cat("\n---------------------------------\n\n")
  cat(paste0("#### Cause of Stop: ", rfs_list[i,2]))
  cat("\n\n")
  cat(paste0("##### Descriptive Data"))
  
  race_data = RIPA_MARIN_2024 |>
    filter(REASON_FOR_STOP == rfs) |>
    mutate(
      race = case_when(
        RAE_FULL == 1 ~ "Asian",
        RAE_FULL == 2 ~ "Black",
        RAE_FULL == 3 ~ "Hispanic",
        RAE_FULL == 4 ~ "MiddleEastern_SouthAsian",
        RAE_FULL == 5 ~ "NativeAmerican",
        RAE_FULL == 6 ~ "PacificIslander",
        RAE_FULL == 7 ~ "White",
        RAE_FULL == 8 ~ "Multiracial",
        TRUE ~ NA_character_
        ),
      result_arrest = if_else(ROS_CUSTODIAL_WITHOUT_WARRANT == 1, 1, 0),
      result_citation = if_else(ROS_IN_FIELD_CITE_RELEASE == 1, 1, 0),
      searched = if_else(NFA_SEARCH_PERSON == 1 | NFA_SEARCH_PROPERTY == 1, 1, 0),
      force_used = if_else(
        OFA_HANDCUFFED == 1 | OFA_PHYSICAL_COMPLIANCE == 1 | OFA_FIREARM_POINT == 1 | 
        OFA_FIREARM_DISCHARGE == 1, 1, 0),
      no_action = if_else(ROS_NO_ACTION == 1, 1, 0)
      )|>
    filter(!is.na(race))
  
  race_summary = race_data |>
    group_by(race) |>
    summarise(
      n = n(),
      total_arrest = sum(result_arrest, na.rm = TRUE),
      arrest_rate = round(mean(result_arrest, na.rm = TRUE)*100, digits = 3),
      total_citation = sum(result_citation, na.rm = TRUE),
      citation_rate = round(mean(result_citation, na.rm = TRUE)*100, digits = 3),
      total_search = sum(searched, na.rm = TRUE),
      search_rate = round(mean(searched, na.rm = TRUE)*100, digits = 3),
      total_force = sum(force_used, na.rm = TRUE),
      force_rate = round(mean(force_used, na.rm = TRUE)*100, digits = 3),
      total_no_action = sum(no_action, na.rm = TRUE),
      no_action_rate = round(mean(no_action, na.rm = TRUE)*100, digits = 3),
      .groups = "drop"
      )|>
    filter(n>=10)
    colnames(c("Population Stoped", "Arrest", "% of Arrest", "Citation", "% of Citation",
               "Searched", "% of Search", "Force", "% of Force", "No Action", "% of No Action"))

  if (nrow(race_summary) > 1) {
    print(kable(race_summary))
    or_arrest = calc_or_ci(race_summary, "total_arrest", "race", "White", "Arrest")
    #or_citation = calc_or_ci(race_summary, "total_citation", "race", "White", "Citation")
    #or_search = calc_or_ci(race_summary, "total_search", "race", "White", "Search")
    #or_force = calc_or_ci(race_summary, "total_force", "race", "White", "Use of Force")
    #or_noaction = calc_or_ci(race_summary, "total_no_action", "race", "White", "No Action")
    or_all = rbind(or_arrest) #, or_citation, or_search, or_force, or_noaction)
    
    cat("\n\n")
    cat(paste0("##### Odds Ratio and 95% Confidence Interval"))
    print(kable(or_all))
    cat("\n\n")
    cat(paste0("##### Statistical Significant Conclusion"))
    conclusion = FALSE
    cat("\n\n")
    
    for (n in 1:dim(or_all)[1]){
      if (or_all[n,3]>1 | is.na(or_all[n,3]) == TRUE){
        cat(paste0("\n- Within ", rfs_list[i,2], " stops, ", or_all[n,1], " individuals experienced ", round(or_all[n,2], digits = 3), " times the odds of ", or_all[n,5], ". The confidence intervals exclude 1.0, indicating statistical significant disparities. \n"))
        conclusion = TRUE
      }
    }
    if (conclusion == FALSE) {
        cat(paste0("\n - Within ", rfs_list[i,2]," stops, no statistically significant disparities were identified for ", or_all[n,1], " individuals.\n"))
    }
    
    } else{
    cat(paste0("\n\n - Less than one sample has a size greater than 10, not signifiant enough to compare.\n\n"))
  }  
}
```

---

## National Origin (using language proficiency as nexuses)

**Methodology**: This analysis uses 2024 California Racial and Identity Profiling Act (RIPA) data from Marin County to examine whether law enforcement outcomes vary based on English language proficiency among individuals stopped for similar reasons. To promote comparability, the analysis is conducted separately for each of the following stop-reason categories (`REASON_FOR_STOP`):

- 1: Traffic Violation
- 2: Reasonable suspicion the person was engaged in criminal activity
- 3: Parole/probation/PRCS/mandatory supervision
- 9: Probable cause to arrest or search

English language proficiency is operationalized using the LIMITED_ENGLISH_FLUENCY variable, recoded into two categories: Limited or no English fluency and Fluent in English. Observations with missing language information are excluded from the analysis.

Each stop is converted into a series of binary indicators representing key enforcement outcomes. 
- Arrests are defined as custodial arrests without a warrant (`ROS_CUSTODIAL_WITHOUT_WARRANT == 1`).
- Citations are defined as in-field cite-and-release dispositions (`ROS_IN_FIELD_CITE_RELEASE == 1`). 
- Searches are defined as any search of a person or property (`NFA_SEARCH_PERSON == 1` or `NFA_SEARCH_PROPERTY == 1`). 
- Use of force is defined as any instance of handcuffing, physical compliance techniques, firearm pointing, or firearm discharge (`OFA_HANDCUFFED`, `OFA_PHYSICAL_COMPLIANCE`, `OFA_FIREARM_POINT`, or `OFA_FIREARM_DISCHARGE`). 
- Stops resulting in no enforcement action are identified by `ROS_NO_ACTION == 1`.

For each stop-reason category, stops are grouped by English proficiency. The analysis calculates the total number of stops, counts of each enforcement outcome, and corresponding rates (means of binary indicators) for each language group.

To maintain statistical stability, language groups with fewer than 10 observations within a stop-reason category are excluded from the summary. Comparative tables are produced only when at least two groups meet this threshold. The resulting summaries provide descriptive measures of enforcement outcomes for individuals with differing English proficiency under comparable stop circumstances, allowing identification of potential disparities in law enforcement actions.

```{r, results = 'asis'}
for (i in 1:dim(rfs_list)[1]) {
  rfs = as.numeric(rfs_list[i,1])
  
  cat("\n---------------------------------\n\n")
  cat(paste0("#### Cause of Stop: ", rfs_list[i,2]))
  cat("\n\n") 
  cat(paste0("##### Descriptive Data"))
  cat("\n\n") 
  
  language_data = RIPA_MARIN_2024 |>
    filter(REASON_FOR_STOP == rfs) |>
    mutate(
      language = case_when(
        LIMITED_ENGLISH_FLUENCY == 1 ~ "Limited or no English fluency",
        LIMITED_ENGLISH_FLUENCY == 0 ~ "Fluent in English",
        TRUE ~ NA_character_
        ),
      result_arrest = if_else(ROS_CUSTODIAL_WITHOUT_WARRANT == 1, 1, 0),
      result_citation = if_else(ROS_IN_FIELD_CITE_RELEASE == 1, 1, 0),
      searched = if_else(NFA_SEARCH_PERSON == 1 | NFA_SEARCH_PROPERTY == 1, 1, 0),
      force_used = if_else(
        OFA_HANDCUFFED == 1 | OFA_PHYSICAL_COMPLIANCE == 1 | OFA_FIREARM_POINT == 1 | 
        OFA_FIREARM_DISCHARGE == 1, 1, 0),
      no_action = if_else(ROS_NO_ACTION == 1, 1, 0)
      ) |>
    filter(!is.na(language))
  
  language_summary = language_data |>
    group_by(language) |>
    summarise(
      n = n(),
      total_arrest = sum(result_arrest, na.rm = TRUE),
      arrest_rate = mean(result_arrest, na.rm = TRUE),
      total_citation = sum(result_citation, na.rm = TRUE),
      citation_rate = mean(result_citation, na.rm = TRUE),
      total_search = sum(searched, na.rm = TRUE),
      search_rate = mean(searched, na.rm = TRUE),
      total_force = sum(force_used, na.rm = TRUE),
      force_rate = mean(force_used, na.rm = TRUE),
      total_no_action = sum(no_action, na.rm = TRUE),
      no_action_rate = mean(no_action, na.rm = TRUE),
      .groups = "drop"
      ) |>
    filter(n >= 10)
      
  if (nrow(language_summary) > 1) {
    print(kable(language_summary))
   or_arrest = calc_or_ci(language_summary, "total_arrest", "language", "Fluent in English", "Arrest")
    #or_citation = calc_or_ci(language_summary, "total_citation", "Fluent in English", "Citation")
    #or_search = calc_or_ci(language_summary, "total_search", "Fluent in English", "Search")
    #or_force = calc_or_ci(language_summary, "total_force", "Fluent in English", "Use of Force")
    #or_noaction = calc_or_ci(language_summary, "total_no_action", "Fluent in English", "No Action")
    or_all = rbind(or_arrest) #, or_citation, or_search, or_force, "Fluent in English", or_noaction)
    
    cat("\n\n")
    cat(paste0("##### Odds Ratio and 95% Confidence Interval"))
    print(kable(or_all))
    cat("\n\n")
    cat(paste0("##### Statistical Significant Conclusion"))
    conclusion = FALSE
    cat("\n\n")
    
    for (n in 1:dim(or_all)[1]){
      if (or_all[n,3]>1 | is.na(or_all[n,3]) == TRUE){
        cat(paste0("\n- Within ", rfs_list[i,2], " stops, ", or_all[n,1], " individuals experienced ", round(or_all[n,2], digits = 3), " times the odds of ", or_all[n,5], ". The confidence intervals exclude 1.0, indicating statistical significant disparities. \n"))
        conclusion = TRUE
      }
    }
      if (conclusion == FALSE) {
        cat(paste0("\n - Within ", rfs_list[i,2]," stops, no statistically significant disparities were identified for ", or_all[n,1], " individuals.\n"))
      }
    } else{
    cat(paste0("\n\n - Less than one sample has a size greater than 10, not signifiant enough to compare.\n\n"))
  }  
}
```

---

# Law Enforcement Specific Data

```{r, results='asis'}

for (i in 1:length(agency_name_list)) {
  agency_name = agency_name_list[i]

    cat("\n---------------------------------\n\n")
    cat(paste0("\n\n## ", agency_name, "\n\n"))
    cat("\n---------------------------------\n\n")
    cat("\n\n") 
    cat(paste0("### Race and Ethnicities"))
    cat("\n\n") 
    

  for (j in 1:dim(rfs_list)[1]) {
    rfs = as.numeric(rfs_list[j,1])
    
    cat("\n---------------------------------\n\n")
    cat(paste0("#### Cause of Stop: ", rfs_list[j,2]))
    cat("\n\n")
    cat(paste0("##### Descriptive Data"))
    cat("\n\n")
    
    race_data = RIPA_MARIN_2024 |>
      filter(REASON_FOR_STOP == rfs & AGENCY_NAME == agency_name) |>
      mutate(
        race = case_when(
          RAE_FULL == 1 ~ "Asian",
          RAE_FULL == 2 ~ "Black",
          RAE_FULL == 3 ~ "Hispanic",
          RAE_FULL == 4 ~ "MiddleEastern_SouthAsian",
          RAE_FULL == 5 ~ "NativeAmerican",
          RAE_FULL == 6 ~ "PacificIslander",
          RAE_FULL == 7 ~ "White",
          RAE_FULL == 8 ~ "Multiracial",
          TRUE ~ NA_character_
          ),
        result_arrest = if_else(ROS_CUSTODIAL_WITHOUT_WARRANT == 1, 1, 0),
        result_citation = if_else(ROS_IN_FIELD_CITE_RELEASE == 1, 1, 0),
        searched = if_else(NFA_SEARCH_PERSON == 1 | NFA_SEARCH_PROPERTY == 1, 1, 0),
        force_used = if_else(
          OFA_HANDCUFFED == 1 | OFA_PHYSICAL_COMPLIANCE == 1 | OFA_FIREARM_POINT == 1 | 
          OFA_FIREARM_DISCHARGE == 1, 1, 0),
        no_action = if_else(ROS_NO_ACTION == 1, 1, 0)
        )|>
      filter(!is.na(race))
    
    race_summary = race_data |>
      group_by(race) |>
      summarise(
        n = n(),
        total_arrest = sum(result_arrest, na.rm = TRUE),
        arrest_rate = round(mean(result_arrest, na.rm = TRUE)*100, digits = 3),
        total_citation = sum(result_citation, na.rm = TRUE),
        citation_rate = round(mean(result_citation, na.rm = TRUE)*100, digits = 3),
        total_search = sum(searched, na.rm = TRUE),
        search_rate = round(mean(searched, na.rm = TRUE)*100, digits = 3),
        total_force = sum(force_used, na.rm = TRUE),
        force_rate = round(mean(force_used, na.rm = TRUE)*100, digits = 3),
        total_no_action = sum(no_action, na.rm = TRUE),
        no_action_rate = round(mean(no_action, na.rm = TRUE)*100, digits = 3),
        .groups = "drop"
        )|>
      filter(n>=10)
      colnames(c("Population Stoped", "Arrest", "% of Arrest", "Citation", "% of Citation", "Searched", "% of Search", "Force", "% of Force", "No Action", "% of No Action"))
  
    if (nrow(race_summary) > 1) {
      print(kable(race_summary))
      or_arrest = calc_or_ci(race_summary, "total_arrest", "race", "White", "Arrest")
      #or_citation = calc_or_ci(race_summary, "total_citation", "race", "White", "Citation")
      #or_search = calc_or_ci(race_summary, "total_search", "race", "White", "Search")
      #or_force = calc_or_ci(race_summary, "total_force", "race", "White", "Use of Force")
      #or_noaction = calc_or_ci(race_summary, "total_no_action", "race", "White", "No Action")
      or_all = rbind(or_arrest) #, or_citation, or_search, or_force, or_noaction)
      
      cat("\n\n")
      cat(paste0("##### Odds Ratio and 95% Confidence Interval"))
      print(kable(or_all))
      cat("\n\n")
      cat(paste0("##### Statistical Significant Conclusion"))
      conclusion = FALSE
      cat("\n\n")
      
      for (n in 1:dim(or_all)[1]){
        if (or_all[n,3]>1 | is.na(or_all[n,3]) == TRUE){
          cat(paste0("\n- Within ", rfs_list[j,2], " stops, ", or_all[n,1], " individuals experienced ", round(or_all[n,2], digits = 3), " times the odds of ", or_all[n,5], ". The confidence intervals exclude 1.0, indicating statistical significant disparities. \n"))
          conclusion = TRUE
        }
      }
      if (conclusion == FALSE) {
          cat(paste0("\n - Within ", rfs_list[j,2]," stops, no statistically significant disparities were identified for ", or_all[n,1], " individuals.\n"))
      }
      
      } else{
      cat(paste0("\n\n - Less than one sample has a size greater than 10, not signifiant enough to compare.\n\n"))
    }  
  }
    
    cat("\n---------------------------------\n\n")
    cat(paste0("### National Origin (using language proficiency as nexuses)"))
    cat("\n\n") 
    
    for (j in 1:dim(rfs_list)[1]) {
    rfs = as.numeric(rfs_list[j,1])
    
    cat("\n---------------------------------\n\n")
    cat(paste0("#### Cause of Stop: ", rfs_list[j,2]))
    cat("\n\n") 
    cat(paste0("##### Descriptive Data"))
    cat("\n\n") 
    
    language_data = RIPA_MARIN_2024 |>
      filter(REASON_FOR_STOP == rfs & AGENCY_NAME == agency_name) |>
      mutate(
        language = case_when(
          LIMITED_ENGLISH_FLUENCY == 1 ~ "Limited or no English fluency",
          LIMITED_ENGLISH_FLUENCY == 0 ~ "Fluent in English",
          TRUE ~ NA_character_
          ),
        result_arrest = if_else(ROS_CUSTODIAL_WITHOUT_WARRANT == 1, 1, 0),
        result_citation = if_else(ROS_IN_FIELD_CITE_RELEASE == 1, 1, 0),
        searched = if_else(NFA_SEARCH_PERSON == 1 | NFA_SEARCH_PROPERTY == 1, 1, 0),
        force_used = if_else(
          OFA_HANDCUFFED == 1 | OFA_PHYSICAL_COMPLIANCE == 1 | OFA_FIREARM_POINT == 1 | 
          OFA_FIREARM_DISCHARGE == 1, 1, 0),
        no_action = if_else(ROS_NO_ACTION == 1, 1, 0)
        ) |>
      filter(!is.na(language))
    
    language_summary = language_data |>
      group_by(language) |>
      summarise(
        n = n(),
        total_arrest = sum(result_arrest, na.rm = TRUE),
        arrest_rate = mean(result_arrest, na.rm = TRUE),
        total_citation = sum(result_citation, na.rm = TRUE),
        citation_rate = mean(result_citation, na.rm = TRUE),
        total_search = sum(searched, na.rm = TRUE),
        search_rate = mean(searched, na.rm = TRUE),
        total_force = sum(force_used, na.rm = TRUE),
        force_rate = mean(force_used, na.rm = TRUE),
        total_no_action = sum(no_action, na.rm = TRUE),
        no_action_rate = mean(no_action, na.rm = TRUE),
        .groups = "drop"
        ) |>
      filter(n >= 10)
        
    if (nrow(language_summary) > 1) {
      print(kable(language_summary))
     or_arrest = calc_or_ci(language_summary, "total_arrest", "language", "Fluent in English", "Arrest")
      #or_citation = calc_or_ci(language_summary, "total_citation", "Fluent in English", "Citation")
      #or_search = calc_or_ci(language_summary, "total_search", "Fluent in English", "Search")
      #or_force = calc_or_ci(language_summary, "total_force", "Fluent in English", "Use of Force")
      #or_noaction = calc_or_ci(language_summary, "total_no_action", "Fluent in English", "No Action")
      or_all = rbind(or_arrest) #, or_citation, or_search, or_force, "Fluent in English", or_noaction)
      
      cat("\n\n")
      cat(paste0("##### Odds Ratio and 95% Confidence Interval"))
      print(kable(or_all))
      cat("\n\n")
      cat(paste0("##### Statistical Significant Conclusion"))
      conclusion = FALSE
      cat("\n\n")
      
      for (n in 1:dim(or_all)[1]){
        if (or_all[n,3]>1 | is.na(or_all[n,3]) == TRUE){
          cat(paste0("\n- Within ", rfs_list[j,2], " stops, ", or_all[n,1], " individuals experienced ", round(or_all[n,2], digits = 3), " times the odds of ", or_all[n,5], ". The confidence intervals exclude 1.0, indicating statistical significant disparities. \n"))
          conclusion = TRUE
        }
      }
        if (conclusion == FALSE) {
          cat(paste0("\n - Within ", rfs_list[j,2]," stops, no statistically significant disparities were identified for ", or_all[n,1], " individuals.\n"))
        }
      } else{
      cat(paste0("\n\n - Less than one sample has a size greater than 10, not signifiant enough to compare.\n\n"))
    }  
  }
}
```

# Conclusion

The Marin County 2024 RIPA stop data identify several statistically significant disparities in arrest outcomes across both racial/ethnic groups and English-language proficiency categories after conditioning on the reason for the stop. These findings compare individuals who were stopped under similar stated circumstances and therefore provide evidence of differential enforcement outcomes that may warrant further examination under the framework of Penal Code § 745(a)(3).
Among racial and ethnic groups, Hispanic individuals experienced significantly higher odds of custodial arrest than White individuals across multiple stop categories. Specifically, Hispanic individuals had 2.568 times the odds of arrest during traffic-violation stops, 1.663 times the odds during reasonable-suspicion stops, and 1.637 times the odds during probable-cause stops. Black individuals experienced 2.231 times the odds of arrest during traffic-violation stops, while Pacific Islander individuals experienced 5.978 times the odds of arrest during reasonable-suspicion stops. No statistically significant racial or ethnic disparities were identified in parole/probation-related stops.

Analysis using limited English proficiency as a proxy for national origin also revealed significant disparities. Individuals perceived to have limited or no English fluency experienced 1.561 times the odds of arrest during reasonable-suspicion stops and 3.968 times the odds of arrest during probable-cause stops compared with individuals perceived to be fluent in English. No statistically significant language-based disparities were identified in traffic-violation or parole/probation-related stops.

Taken together, the results show that statistically significant disparities in arrest outcomes are concentrated in traffic-violation, reasonable-suspicion, and probable-cause stops, while parole/probation-related stops did not demonstrate significant differences. The most consistent pattern observed was the elevated likelihood of arrest among Hispanic individuals across multiple stop categories and among individuals perceived to have limited English proficiency during criminal-investigation-related stops. Because arrests are often the initiating event that exposes individuals to prosecution and potential conviction, these disparities may be relevant to assessing whether similarly situated individuals are treated differently on the basis of race, ethnicity, or national origin.

Although these findings do not establish causation or independently prove a violation of the California Racial Justice Act, they constitute aggregate statistical evidence of differential treatment that may support further inquiry. Additional analysis incorporating offense-specific conduct, agency-level practices, charging decisions, and conviction outcomes would be necessary to determine whether the observed arrest disparities translate into prosecutorial disparities contemplated by Penal Code § 745(a)(3).
